---
title: "Paediatric Mental Health Attendances - CMAST"
author: "NHS Transformation Unit"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
 html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    toc_collapsed: true
    css: ../config/nhs_tu_theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, out.width = '100%')
library(here)

source(paste0(here(), "/src/requirements/packages.R"))
source(paste0(here(), "/src/config/palette.R"))
source(paste0(here(), "/src/config/tu_ggplot_themes.R"))
source(paste0(here(), "/src/load/load.R"))
source(paste0(here(), "/src/processing/processing.R"))
source(paste0(here(), "/src/visualising/system_level.R"))

```

## Introduction
***

This analysis has been undertaken to understand recent trends and changes in the volume and waiting times for paediatric patients attending the Emergency Department (ED) due to mental health reasons across CMAST providers. The analysis uses the [Emergency Care Dataset](https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/emergency-care-data-set-ecds) (ECDS) as the data source and the national definition of mental health attendances in ED was used to identify relevant activity. The analysis looks back over activity since April 2019 through to December 2023. More information on the methods used can be found in the [Methodology](#Methodology) section.

This report contains the following analysis:

* __Mental health attendances as ED__: Analysis into the number of attendances across CMAST, individual providers and at site level. This is further broken down to explore when presentations occur by day of the week and hour of the day.

* __Waiting times in ED__: Analysis into how long patients are waiting in the department across CMAST, individual providers and at site level.

* __Outcomes from ED__: Analysis into where patients go to when they leave the department, such as admission rates.

* __Data Quality__: Exploration of the data quality of ECDS used in this analysis.

<br/>

## Attendances at ED
***
This section explores how attendances at the ED for mental health reasons have been changing since pre-pandemic.

At the CMAST level there is very little variation across the time period. There is however notable variation when comparing trusts and individual sites.

### Attendance volumes {.tabset .tabset-fade}

#### CMAST

The plot below shows the number of attendances to the ED each month at CMAST providers.
```{r activity_system_ts, dpi = 1000}

plot_act_ts_system

```


<br/>

#### Provider

<br/>

#### Site Level

<br/>

### Arrival patterns - out of hours {.tabset .tabset-fade}

#### CMAST

The plot below shows the number of attendances to the ED each month at CMAST providers. With the exception of the beginning of the COVID19 pandemic and some summer months, attendances and split almost half way between in hours (08:00 to 18:00) and out of hours (18:00 to 08:00).
```{r activity_system_io, dpi = 1000}

plot_act_system_in_out

```


<br/>

#### Provider

<br/>

#### Site Level

<br/>

### Arrival patterns - arrival hour {.tabset .tabset-fade}

#### CMAST

The plot below shows the number of attendances to the ED of CMAST providers depending on the day of the week and the hour of the day. Number of attendances follows a similar pattern for week days, with general decreases in attendances between midnight (hour 0) and 06:00, with a sharp increase between 08:00 and 10:00. After this it remains high and consistent until the approach to 20:00 where a decrease starts. Weekends see gentler trends, with Sundays showing a peak in attendances at approximately 22:00 and a low at 06:00, with almost smooth transitions between.
```{r activity_system_dh, dpi = 1000}

plot_act_profile_dh

```


<br/>

#### Provider

<br/>

#### Site Level

<br/>

## Waiting times in Emergency Department
***

This section provides an overview of how long patients waited in the Emergency Department from their arrival to their departure.

### Waiting times pattern {.tabset .tabset-fade}

#### CMAST

The plot below shows the median time spent in the department, from arrival to departure, accompanied by the interquartile range and the 10th to 90th percentile range. The median wait time was almost always below the 4 hour A&E target, with a significant waiting time reduction from the start of the COVID19 pandemic. This started to increase and reached pre pandemic times around Jan-2021. From Apr-2021 the interquartile range and 10th to 90th Percentile range shows a significant increase, with the upper quartile boundary exceeding the 4 hour target by 2.5 hours in Oct-2021, exceeding pre pandemic highs. Median wait times, along with ranges, have seen very little change since mid 2021, with median wait times now consistently breaching the 4 hour A&E target.
```{r wait_time_system, dpi = 1000}

plot_act_wait_ts

```

<br/>

#### Provider

<br/>

#### Site

<br/>

## Onward destination following Emergency Department
***

This section provides an overview of where patients went following their discharge from ED.

### ED outcome pattern {.tabset .tabset-fade}

#### CMAST

The plot below shows the outcome of patients attending ED and the onward destination, whether that be an admission, discharge or transfer to an alternate provider. Patient outcomes have not seen much change at the system level across the period, with the exception of reduced numbers following the start of the COVID19 pandemic. Most patients are discharged from the hospital, whether that be to their home or other place of residence or into police custody. Since Apr-2021 there have been more transfers of care per month between providers following an attendance at ED, though a significant proportion are still admitted to the hospital of presentation following an ED attendance.
```{r outcome_system, dpi = 1000}

plot_act_system_outcomes_ts

```

<br/>

#### Provider

<br/>

#### Site

<br/>

### ED outcome by day and by time {.tabset .tabset-fade}

#### CMAST

The plot below shows the outcome depending on the day of the week and the hour of the day.
```{r outcome_system_day, dpi = 1000}

plot_outcome_profile_d

```

```{r outcome_system_hour, dpi = 1000}

plot_outcome_profile_h

```

<br/>

#### Provider

<br/>

#### Site

<br/>

## Methodology {#Methodology}

### Data Sources
This analysis has been undertaken on the [Emergency Care Dataset](https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/emergency-care-data-set-ecds) (ECDS).

### Definition of Mental Health Attendances
This analysis has defined attendances at the Emergency Department as _mental health related_ using the national definition from NHS England. This definition has been defined by the Mental Health Analytics team at NHS England with the information on the codes used [here](https://github.com/nhsengland/MHSDS/blob/f649d73d5573b4d3ab63215bb15fedd5c195e4c9/UEMH%20dashboard/ECDS%20Crisis#L66). This definition flags an attendance as a _mental health_ attendance when either the **Chief Complaint Code**, **Diagnosis** or the **Injury Intent Code** is one of the recognised codes. For the purposes of this analysis _Paediatric attendances_ are those where the patient was 17 years or under at the date of their arrival to the Emergency Department.

### Analysis
This analysis has been undertaken in the open-source programming language R. The codebase for this analysis is available to view on its dedicated [Github repository](https://github.com/NHS-Transformation-Unit/CMAST).