---
title: "Paediatric Mental Health Attendances - CMAST"
author: "NHS Transformation Unit"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
 html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    toc_collapsed: true
    css: ../config/nhs_tu_theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, out.width = '100%')
library(here)

source(paste0(here(), "/src/requirements/packages.R"))
source(paste0(here(), "/src/config/palette.R"))
source(paste0(here(), "/src/config/tu_ggplot_themes.R"))
source(paste0(here(), "/src/load/load.R"))
source(paste0(here(), "/src/processing/processing.R"))
source(paste0(here(), "/src/visualising/system_level.R"))

```

## Introduction
***

This analysis has been undertaken to understand recent trends and changes in the volume and waiting times for paediatric patients attending the Emergency Department (ED) due to mental health reasons across CMAST providers. The analysis uses the [Emergency Care Dataset](https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/emergency-care-data-set-ecds) (ECDS) as the data source and the national definition of mental health attendances in ED was used to identify relevant activity. The analysis looks back over activity since April 2019 through to December 2023. More information on the methods used can be found in the [Methodology](#Methodology) section.

This report contains the following analysis:

* __Mental health attendances as ED__: Analysis into the number of attendances across CMAST, individual providers and at site level. This is further broken down to explore when presentations occur by day of the week and hour of the day.

* __Waiting times in ED__: Analysis into how long patients are waiting in the department across CMAST, individual providers and at site level.

* __Outcomes from ED__: Analysis into where patients go to when they leave the department, such as admission rates.

* __Data Quality__: Exploration of the data quality of ECDS used in this analysis.

<br/>

## Attendances at ED
***
This section explores how attendances at the ED for mental health reasons have been changing since pre-pandemic.

### Attendance volumes {.tabset .tabset-fade}

#### CMAST

The plot below shows the number of attendances to the ED each month at CMAST providers.
```{r activity_system_ts, dpi = 1000}

plot_act_ts_system

```


<br/>

#### Provider

<br/>

#### Site Level

<br/>

### Arrival patterns - out of hours {.tabset .tabset-fade}

#### CMAST

<br/>

#### Provider

<br/>

#### Site Level

<br/>

### Arrival patterns - arrival hour {.tabset .tabset-fade}

#### CMAST

<br/>

#### Provider

<br/>

#### Site Level

<br/>

## Waiting times in Emergency Department
***

This section provides an overview of how long patients waited in the Emergency Department from their arrival to their departure.

<br/>

## Methodology {#Methodology}

### Data Sources
This analysis has been undertaken on the [Emergency Care Dataset](https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/emergency-care-data-set-ecds) (ECDS).

### Definition of Mental Health Attendances
This analysis has defined attendances at the Emergency Department as _mental health related_ using the national definition from NHS England. This definition has been defined by the Mental Health Analytics team at NHS England with the information on the codes used [here](https://github.com/nhsengland/MHSDS/blob/f649d73d5573b4d3ab63215bb15fedd5c195e4c9/UEMH%20dashboard/ECDS%20Crisis#L66). This definition flags an attendance as a _mental health_ attendance when either the **Chief Complaint Code**, **Diagnosis** or the **Injury Intent Code** is one of the recognised codes. For the purposes of this analysis _Paediatric attendances_ are those where the patient was 17 years or under at the date of their arrival to the Emergency Department.

### Analysis
This analysis has been undertaken in the open-source programming language R. The codebase for this analysis is available to view on its dedicated [Github repository](https://github.com/NHS-Transformation-Unit/CMAST).