---
title: |
  | Paediatric Mental Health Attendances
  | CMAST
author: "<a href='https://github.com/ASW-Analyst'>Andy Wilson</a><br><a href='https://github.com/SiWickham'>Simon Wickham</a><br><a href='https://transformationunit.nhs.uk/'>NHS Transformation Unit</a>"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
 html_document:
    toc: true
    toc_depth: 2
    toc_float: true
    toc_collapsed: true
    css: ../config/nhs_tu_theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, out.width = '100%')
library(here)

source(paste0(here(), "/src/requirements/packages.R"))
source(paste0(here(), "/src/config/palette.R"))
source(paste0(here(), "/src/config/tu_ggplot_themes.R"))
source(paste0(here(), "/src/load/load.R"))
source(paste0(here(), "/src/processing/processing.R"))
source(paste0(here(), "/src/processing/dq.R"))
source(paste0(here(), "/src/visualising/system_level.R"))
source(paste0(here(), "/src/visualising/trust_level.R"))
source(paste0(here(), "/src/outputs/text.R"))

```

```{r logo, echo = FALSE}

htmltools::img(src = knitr::image_uri(paste0(here(), "/images/TU_logo_large.png")), 
               alt = 'logo', 
               style = 'position:absolute; top:0; right:5%; padding:10px;',
               width = "180px",
               heigth = "180px")

```


## Introduction
***

This analysis has been undertaken to understand recent trends and changes in the volume and waiting times for paediatric patients attending the Emergency Department (ED) due to mental health reasons across CMAST providers. The analysis uses the [Emergency Care Dataset](https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/emergency-care-data-set-ecds) (ECDS) as the data source and the national definition of mental health attendances in ED was used to identify relevant activity. The analysis looks back over activity since April 2019 through to December 2023. More information on the methods used can be found in the [Methodology](#Methodology) section.

This report contains the following analysis:

* __Mental health attendances as ED__: Analysis into the number of attendances across CMAST and constituent providers. This is further broken down to explore when patients present to the Emergency Department across the week.

* __Waiting times in ED__: Analysis into how long patients are waiting in the department across CMAST and constituent providers.

* __Outcomes from ED__: Analysis into where patients go to when they leave the department, such as admission rates.

* __Data Quality__: Exploration of the data quality of ECDS used in this analysis.

<br/>

## Attendances at ED
***
This section explores the volume of attendances to Emergency Departments and how these have changed over. The timing of presentations to the ED across the hours of the day is also included.

### Attendance volumes {.tabset .tabset-fade}

#### CMAST

The plot below shows the number of attendances to the ED each month at CMAST providers. There is significant variation in attendances across the time period, however their is no discernable trend in the number of attendances each month. The latest 12 month rolling average has a figure of `r attends_system_rolling_latest[[1, 8]]` which has changed little in recent months.

```{r activity_system_ts, dpi = 1000}

plot_act_ts_system

```


<br/>

#### Provider

The below plot shows the number of attendances to the ED each month by CMAST providers. There is significant variation across these sites. Data at Liverpool University Hospitals doesn't identify any Mental Health attendances until late in 2019, with a subsequent increase again in Apr-2021. Mersey and West Lancashire Hospitals also sees an increase during the time period. Conversely, Wirral University Hospital sees a general decrease in attendances over the time period and Alder Hey shows a significant drop in attendances from Apr-2023.
```{r activity_trust_ts, dpi = 1000}

plot_act_ts_trust

```

<br/>

### Arrivals to ED - Attendances out-of-hours {.tabset .tabset-fade}
For the purposes of this analysis _out-of-hours_ attendances are defined as those where the patient arrived at the ED after 18:00 and before 08:00.

#### CMAST
The chart below shows the proportion of attendances each month that occurred _out-of-hours_. Whilst there was a slight increase in the proportion of attendances arriving out-of-hours during the early months of the pandemic, there doesn't appear to be any discernible trend. There is significant month-on-month variation in the proportion of attendances arriving out-of-hours.

```{r activity_system_o, dpi = 1000}

plot_act_system_out

```

<br/>

#### Provider

The chart below shows the variation of out-of-hours attendances as a proportion of all attendances for each of the providers in CMAST. This volatility in the proportion attending out-of-hours across CMAST can be seen repeated for each provider. There does appear to be an increasing trend at Liverpool University Hospitals, with a higher proportion now attending out-of-hours compared with pre-pandemic.

```{r activity_trust_pout, dpi = 1000}

plot_act_trust_pout

```

<br/>


### Arrivals to ED - Arrival Times {.tabset .tabset-fade}

We have also explored the profile of these attendances across the days of the week and the hours of the day to understand when people present. This analysis has been undertaken on data for the whole period from April 2019 - December 2023.

#### CMAST

The chart below shows the number of attendances to the ED of CMAST providers by each hour of the day, across each day.The weekdays follow a similar pattern with fewer people attending between the hours of 06:00 and 08:00 and then an increase across the late morning. Attendance numbers then remain high throughout the afternoon and into the evening. The weekends have a slightly different pattern with attendances rising across the daytime and peaking in the late evening. 

```{r activity_system_dh, dpi = 1000}

plot_act_profile_dh

```

<br/>

#### Provider {.tabset .tabset-fade}

The CMAST level analysis has been repeated for each of the constituent providers. Select the tab below to see the arrival profile for that provider.

##### Alder Hey
```{r activity_trust_dh_rbs, dpi = 1000}

plot_act_trust_dh_profile("RBS")

```

<br/>


##### Countess of Chester
```{r activity_trust_dh_rjr, dpi = 1000}

plot_act_trust_dh_profile("RJR")

```

<br/>

##### East Cheshire
```{r activity_trust_dh_rjn, dpi = 1000}

plot_act_trust_dh_profile("RJN")

```

<br/>

##### Liverpool University
```{r activity_trust_dh_rem, dpi = 1000}

plot_act_trust_dh_profile("REM")

```

<br/>

##### Mersey and West Lancashire
```{r activity_trust_dh_rbn, dpi = 1000}

plot_act_trust_dh_profile("RBN")

```

<br/>

##### Mid Cheshire
```{r activity_trust_dh_rbt, dpi = 1000}

plot_act_trust_dh_profile("RBT")

```

<br/>

##### Warrington and Halton
```{r activity_trust_dh_rww, dpi = 1000}

plot_act_trust_dh_profile("RWW")

```

<br/>

##### Wirral University
```{r activity_trust_dh_rbl, dpi = 1000}

plot_act_trust_dh_profile("RBL")

```

<br/>


## Waiting times in Emergency Department
***

This section provides an overview of how long patients waited in the Emergency Department from their arrival to their departure.

### Waiting times pattern {.tabset .tabset-fade}

#### CMAST

The plot below shows the median time spent in the department, from arrival to departure, accompanied by the interquartile range and the 10th to 90th percentile range. The median wait time was almost always below the 4 hour A&E target, with a significant waiting time reduction from the start of the COVID19 pandemic. This started to increase and reached pre pandemic times around Jan-2021. From Apr-2021 the interquartile range and 10th to 90th Percentile range shows a significant increase, with the upper quartile boundary exceeding the 4 hour target by 2.5 hours in Oct-2021, exceeding pre pandemic highs. Median wait times, along with ranges, have seen very little change since mid 2021, with median wait times now consistently breaching the 4 hour A&E target.
```{r wait_time_system, dpi = 1000}

plot_act_wait_ts

```

<br/>

The chart below shows the same waiting time distributions as the previous chart but split between attendances out-of-hours (left-hand chart) and in-hours (right-hand chart). Whilst this shows increase in longer waiting times for both groups, the rise in longer waiting times is more noticeable for patients who attend out-of-hours.

```{r wait_time_system_comp_inout, dpi = 1000}

grid.arrange(plot_act_wait_out_ts, plot_act_wait_in_ts, ncol = 2, top = "Comparison of Waiting Times in ED")

```

<br/>

#### Provider
The charts below show the distribution of waiting times for across each of the constituent providers. There are further charts then splitting these distributions for out-of-hours and in-hours.

```{r wait_time_trust_overall, dpi = 1000}

plot_act_wait_trust_ts

```

<br/>

```{r wait_time_trust_ooh, dpi = 1000}

plot_act_wait_out_trust_ts

```

<br/>

```{r wait_time_trust_inh, dpi = 1000}

plot_act_wait_in_trust_ts

```

<br/>


## Onward destination following Emergency Department
***

In this section we explore the outcome of patients attendances at the Emergency Department. This includes a breakdown of where the patient went to following their stay and the admission rates across CMAST.

### ED Outcome Destinations {.tabset .tabset-fade}

#### CMAST

The chart below shows the outcome destination for patients attending ED. These have been grouped into categories covering discharge, admission, transfer or unknown (i.e. missing data). This chart shows that there does appear to be some coding changes since pre-pandemic with a greater use of **Provider Transfer** codes being used. This potentially indicates that there is better recording of when patients are transferred to another provider, most likely a Mental Health provider, at the end of their stay in the ED.

```{r outcome_system, dpi = 1000}

plot_act_system_outcomes_ts

```

<br/>

The chart below shows the percentage of attendances that were admitted at the end of their stay in the ED. This _should_ represent those patients who are admitted to the acute provider at the end of the stay. This could indicate that these patients are admitted into acute beds.There is appears to show a declining trend in the proportion of attendances that result in an admission. This could be an artifact in the data relating now recorded as transferred to another provider. Another explanation could be due increased bed pressures seen recently meaning their is reduced bed capacity to admit patients to the acute provider.

```{r outcome_system_admit, dpi = 1000}

plot_act_system_outcomes_ts_admit

```

<br/>

#### Provider
The chart below provides a breakdown of the admission rates across each of the constituent providers.

```{r outcome_trust_admit, dpi = 1000}

plot_act_system_outcomes_ts_admit

```

<br/>

## Data Quality
***

As Mental Health attendances are being identified via specific SNOMED codes within the Chief Complaint, Diagnoses and Injury Intent fields within ECDS it is important to consider how many records are missing these codes. This will give an indication of potential Mental Health attendances that could not be identified. The table below gives a breakdown for the period of April 2019 - December 2023 for each of the providers the number of ED attendances for patients aged 17 or under. These are then classified as Mental Health (as per the national definition), physical health or missing. The missing group represent those attendances that do not have a Chief Complaint, Injury Intent or Diagnosis.

```{r pro_dq}

ecds_dq_pro %>%
  rename("Mental Health" = 7,
         "Provider Code" = 1,
         "Provider Name" = 2,) %>%
  mutate(Missing_Rate_format = round(Missing_Rate * 100,2)) %>%
  rename("Missing Rate (%)" = 11) %>%
  select(c(1,2,7,8,9,11)) %>%
  kable(format = "html", align = "llrrrr") %>%
  kable_styling() %>%
  row_spec(0, background = palette_tu[1], color = "white") %>%
  scroll_box(height = "400px")

```

<br/>

## Methodology {#Methodology}
***

### Data Sources
This analysis has been undertaken on the [Emergency Care Dataset](https://digital.nhs.uk/data-and-information/data-collections-and-data-sets/data-sets/emergency-care-data-set-ecds) (ECDS).

### Definition of Mental Health Attendances
This analysis has defined attendances at the Emergency Department as _mental health related_ using the national definition from NHS England. This definition has been defined by the Mental Health Analytics team at NHS England with the information on the codes used [here](https://github.com/nhsengland/MHSDS/blob/f649d73d5573b4d3ab63215bb15fedd5c195e4c9/UEMH%20dashboard/ECDS%20Crisis#L66). This definition flags an attendance as a _mental health_ attendance when either the **Chief Complaint Code**, **Diagnosis** or the **Injury Intent Code** is one of the recognised codes. For the purposes of this analysis _Paediatric attendances_ are those where the patient was 17 years or under at the date of their arrival to the Emergency Department.

### Analysis
This analysis has been undertaken in the open-source programming language R. The codebase for this analysis is available to view on its dedicated [Github repository](https://github.com/NHS-Transformation-Unit/CMAST).